---
title: "Cytokine Loops Part 2-3: 1h vs 20h comparison"
author: "JT"
date: "2026-03-06"
format: 
  pdf: 
    mainfont: "arial"
    monofont: "monaco"
    monofontoptions: 
    - Scale=0.82 # value for 80 character width with monaco font on A4
    code-block-bg: e5e6e4
---

------------------------------------------------------------------------

# Combining analyses from EXP-26-EI5039 and EXP-26EI5040

------------------------------------------------------------------------

{{< pagebreak >}}

## Experiment setup

The two experiments had identical setups, only differing in stimulation time: short and long term incubation.

Blood samples were prepared as following:

75 µL Blood + 75 µL WB-Stim –\> 150 µL cultures + 20 µL pre-treatments+stim added

| Setting | Parameter | EXP-26-EI5039 | EXP-26-EI5040 |
|-----------------|-----------------|-------------------|-------------------|
| Readout | IFNa ELISA | 2 Plates | 2 Plates |
|  | TNF ELISA | 2 Plates | 2 Plates |
| Processing | Hamilton Method | M6 Plasma Harvest v0.3 | M6 Plasma Harvest v0.3 |
| Material | Plasma | 70 µL Plasma first round, 50 µL in second round. | 70 µL Plasma first round, 50 µL in second round. |
| Stimulation | Ctrl | 0 | 0 |
|  | R848 | 500 ng/mL | 500 ng/mL |
|  | LPS | 1 ng/mL | 1 ng/mL |
| Pre-treatment | IFNa | 0, 10 ng/mL | 0, 10 ng/mL |
|  | IFNβ | 0, 10 ng/mL | 0, 10 ng/mL |
|  | IFNγ | 0, 10 ng/mL | 0, 10 ng/mL |
|  | TNF | 0, 10 ng/mL | 0, 10 ng/mL |
|  | IL-1β | 0, 10 ng/mL | 0, 10 ng/mL |
|  | IL-6 | 0, 10 ng/mL | 0, 10 ng/mL |
|  | IL-23 | 0, 10 ng/mL | 0, 10 ng/mL |
| Timepoints | Pre-treatment | 1h | 1h |
|  | Stimulation | 20h | 1h |

: Experimental setup of EXP-26-EI5039 and EXP-26-EI5040 in comparison.

{{< pagebreak >}}

## Set up code environment

### Load libraries

```{r install packages}
# Install packages if not already installed / link from renv cache
required_packages <- c("tidyverse", "conflicted", "drc", "RColorBrewer", 
                       "ggridges", "scales", "colorspace", "Cairo", "patchwork")
for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE)) {
    renv::install(pkg, dependencies = TRUE)
  }
}
```


```{r load packages}
# Imports for curve fitting and data manipulation
library(tidyverse) # for data manipulation and plotting, must-have
library(conflicted) # for managing function conflicts
library(drc) # for standard curve fitting
library(RColorBrewer) # for color palettes
# library(ggridges) # for ridge plots of OD distributions
library(scales) # for scaling functions in plots
# library(colorspace) # for colour manipulation
library(Cairo) # for high-quality PDF output
library(patchwork) # for combining multiple ggplots into one figure
conflicts_prefer(dplyr::filter)
conflict_prefer("select", "dplyr")
```

### Information for reproducibility
```{r reproducibility}
# Output folder
experiment_id = "EXP-26-EI5039-40"
output_folder = paste0("results/", experiment_id, "/")

set.seed(23164) # for reproducing jitter in plots

# Export session info into a txt file
## Source - https://stackoverflow.com/a/21967272
## Posted by Roland
## Retrieved 2026-03-06, License - CC BY-SA 3.0
writeLines(capture.output(sessionInfo()), 
           paste0(output_folder, experiment_id, "-sessionInfo.txt"))
```

{{< pagebreak >}}


## Define global variables and functions

### Varibales
```{r define variables}
# Experiment-related variables
# Identical to earlier run, with same experiment ID and variables. Previous 
# experiment was saved separately.
plate_ids = c("P1", "P2")
cytokines = c("IFNa", "TNF")
# Each plate is a replicate of the same sample set and cytokine.

```

### Functions
```{r functions}
# File paths return function - based on experiment id
get_data_file_path <- function(experiment_id, cytokine, plate_id) {
  paste0(data_file_folder, experiment_id, "_", cytokine, "-", plate_id, ".csv")
}
# Robust csv reader and data importer for general use
read_data_file2 <- function(file_path) {
  data <- read.csv(file_path, header = TRUE, sep = ",", dec = ".", 
                   stringsAsFactors = FALSE)
  return(data)
}
# Data imports
## Data processing and formatting
### Column name adjustment
plate_reader_colname_adjustment <- function(data) {
  # The colnames are always in the format: "Well", "Content", 
  # "Well.Scan..Average..450.1", 
  # "Well.Scan..Average..570.2"/"Well.Scan..Average..620.2")
  # Since we have different reference wavelengths (570 or 620), we rename the 
  # columns to primary and reference OD
  colnames(data) <- gsub("Well.Scan..Average..450.1.", 
                         "OD_primary", 
                         colnames(data))
  if ("Well.Scan..Average..570.2." %in% colnames(data)) {
    colnames(data) <- gsub("Well.Scan..Average..570.2.", 
                           "OD_ref", 
                           colnames(data))
  } else if ("Well.Scan..Average..620.2." %in% colnames(data)) {
    colnames(data) <- gsub("Well.Scan..Average..620.2.", 
                           "OD_ref", 
                           colnames(data))
  }
  selected_cols <- c("Well", "Content", "OD_primary", "OD_ref")
  data <- data[, selected_cols]
  return(data)
}

### OD correction: OD = OD450 - OD570 (alt. OD = OD450 - OD620)
OD_correction <- function(data) {
  data <- data %>%
    mutate(OD_cor = OD_primary - OD_ref)
  return(data)
}
### OD correction: OD = OD450 - OD570 (alt. OD = OD450 - OD620)
# OD_correction <- function(data) {
#   if ("OD570" %in% colnames(data)) {
#     data <- data %>%
#       mutate(OD_cor = OD450 - OD570)
#   } else if ("OD620" %in% colnames(data)) {
#     data <- data %>%
#       mutate(OD_cor = OD450 - OD620)
#   }
#   return(data)
# }

## Mapping file import
### Not needed, using read_data_file function

## Merging into main data frame
### Not needed, using merge function in main pipeline

# Blank correction and sensitivity calculation
## Mean blank calculation
calculate_mean_blank <- function(
    data # data frame with OD values and Content column indicating blanks
    ) {
  # Select blank wells called "Blank B" or "Blank" in Content
  blanks_data = data %>%
    filter(Content == "Blank B" | Content == "Blank")
  mean_blank_od <- blanks_data %>%
    # Calculate mean and sd of blanks
    summarise(mean_blank_OD = mean(OD_cor, na.rm = TRUE)) %>%
    pull(mean_blank_OD)
  sd_blank_od <- blanks_data %>%
    summarise(sd_blank_OD = sd(OD_cor, na.rm = TRUE)) %>%
    pull(sd_blank_OD)
  blank_stats <- data.frame("mean_blank_OD" = mean_blank_od,
                            "sd_blank_OD" = sd_blank_od) 
  return(blank_stats)
}
## Blank subtraction
perform_blank_correction <- function(
    data, blank_stats
    ) {
  data <- data %>%
    mutate(OD_blank_cor = OD_cor - blank_stats$mean_blank_OD) %>%
    # Round to 3 decimal places
    mutate(OD_blank_cor = round(OD_blank_cor, 3)) %>%
    # Set negative values to zero
    mutate(OD_blank_cor = ifelse(OD_blank_cor < 0, 0, OD_blank_cor))
  return(data)
}
## Sensitivity calculation: ULOQ, LLOQ, LLOD
calculate_sensitivity <- function(
    standards_data, blanks_data_frame, cytokine, plate_id
    ) {
  # Retrieve mean and SD of blank
  mean_blank_od <- blanks_data_frame %>%
    filter(ExperimentID == experiment_id,
           Cytokine == cytokine,
           PlateID == plate_id) %>%
    pull(mean_blank_OD)
  sd_blank_od <- blanks_data_frame %>%
    filter(ExperimentID == experiment_id,
           Cytokine == cytokine,
           PlateID == plate_id) %>%
    pull(sd_blank_OD)
  # LLOD = mean blank + 3*SD blank
  LLOD_od <- mean_blank_od + 3 * sd_blank_od
  # LLOQ = mean blank + 10*SD blank
  LLOQ_od <- mean_blank_od + 10 * sd_blank_od
  # ULOQ = Highest standard OD
  ULOQ_od <- max(standards_data$OD_blank_cor, na.rm = TRUE)
  sensitivity_df <- data.frame("Cytokine" = cytokine,
                               "PlateID" = plate_id,
                               "LLOD_OD" = LLOD_od,
                               "LLOQ_OD" = LLOQ_od,
                               "ULOQ_OD" = ULOQ_od)
  return(sensitivity_df)
}

# Standard curve fitting
## Curve fitting - 4PLM, Response = OD, Predictor = Concentration
## Using drm function from drc package by Christian Ritz and Jens C. Streibig
fit_standard_curve <- function(
    standards_data # df with OD_blank_cor and Concentration_pg_mL columns
    ) {
  model <- drm(OD_blank_cor ~ Concentration_pg_mL, 
               data = standards_data, 
               fct = LL.4(names=c("Slope", "Lower", "Upper", "ED50")))
  return(
    model # model object used for predictions and plotting
    )
}

## Concentration calculation
calculate_concentration <- function(
    model, od_values # calculate OD based on model
    ) {
  concentrations <- ED(model, type = "absolute", od_values, display = F)
  return(
    concentrations # 
    )
}

```

## Import data

### Read in data from two experiments

```{r}
# Define experiment IDs for both experiments
experiment_id_20h <- "EXP-26-EI5039"
experiment_id_1h <- "EXP-26-EI5040"
```

Looping over experiment IDs for importing finalised data sets for each experiment, stored as csv files. We have one data set per experiment, with the final results data.

```{r}
# Loop over experiment IDs
experiment_ids <- c(experiment_id_1h, experiment_id_20h)
# Empty data frame to store combined data
combined_data <- data.frame()
for (import_experiment_id in experiment_ids) {
  input_folder = paste0("results/", import_experiment_id, "/")
  file_path <- paste0(input_folder, import_experiment_id, "Output.csv")
  # Read data file
  experiment_data <- read_data_file2(file_path)
  # Add experiment ID column to data frame
  # experiment_data$ExperimentID <- experiment_id # Not needed, already in df
  if ("Method" %in% colnames(experiment_data)) {
    experiment_data <- experiment_data %>%
      select(-Method) # Remove Method column as it is not in EXP-26-EI5040
  }
  # Append to combined data frame
  combined_data <- rbind(combined_data, experiment_data)
}

```

## Plotting

### Sample Concentration Plot

Looking at all samples, and how the OD values fell within range. Above, we labelled each sample with regard to its OD as "Below LLOD", "Between LLOD and LLOQ", "Within Range", or "Above ULOQ". We can use this information to color the points in the plot accordingly.

```{r data frame for plotting}
# Adjust the samples data frame for plotting
samples_data_plot <- combined_data %>%
  select(Well, PlateID, Cytokine, Donor_ID, Sample, PreTreatment, Stim, 
         Stim_Conc, Timepoint, Final_Concentration_pg_mL, Within_Range)

# Sort sample names in alphabetical order
samples_data_plot$Sample <- factor(
  samples_data_plot$Sample, levels = sort(unique(samples_data_plot$Sample))
  )

# Set NA values as 0
samples_data_plot <- samples_data_plot %>%
  mutate(
    Final_Concentration_pg_mL = ifelse(is.na(Final_Concentration_pg_mL), 
                                       0, 
                                       Final_Concentration_pg_mL),
    Sample = as.factor(Sample)
    )

# Make a character vector of sample IDs in the order they appear in the data frame
samples_data_plot$Sample <- as.character(samples_data_plot$Sample)

# Then we can use this vector to set the levels of the Sample factor
samples_data_plot$Sample <- factor(samples_data_plot$Sample, 
                                   levels = unique(samples_data_plot$Sample))

```

```{r sample concentration plot}
# Colours for within range, above ULOQ, below LLOD
range_palette <- c("Below LLOD" = "yellow3", 
                   "Between LLOD and LLOQ" = "grey80", 
                   "Within Range" = "black", 
                   "Above ULOQ" = "red")

# Plot sample concentrations
sample_conc_overview <- ggplot(samples_data_plot, 
                               aes(x = Sample, 
                                   y = Final_Concentration_pg_mL, 
                                   color = Within_Range, 
                                   shape = Donor_ID)) +
  geom_point(size = 3) +
  labs(title = paste0("Plasma concentration of cytokines, 
                      all samples and all ELISAs together"),
       x = "Sample ID",
       y = "Concentration (pg/mL)") +
  theme_minimal() +
  scale_color_manual(values = range_palette) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ Cytokine, ncol = 2, scales = "free_y", axes = "all")

print(sample_conc_overview, width = 12, height = 6)

ggsave(sample_conc_overview, 
       filename = paste0(output_folder, 
                         experiment_id, 
                         "_Sample_conc_overview.pdf"), 
       width = 8, 
       height = 6)
```

### Box Plots

Box plots with stim concentration on x axis, untreated / TNF-treated side-by-side, and final concentration on y axis.

```{r prepare data for boxplots}

# Remove IFNa+TNF samples, as we cannot interpret them for IFNa or TNF ELISA
samples_data_plot <- samples_data_plot %>%
  filter(PreTreatment != "IFNa+TNF")

# Reorder PreTreatment factor levels
samples_data_plot <- samples_data_plot %>%
  mutate(PreTreatment = factor(PreTreatment, 
                               levels = c("Untreated", "IFNa", "IFNb", "IFNg", 
                                          "TNF", "IL-1b", "IL-6", "IL-23")))

# Reorder Stim factor levels
samples_data_plot <- samples_data_plot %>%
  mutate(Stim = factor(Stim, levels = c("Unstim", "LPS", "R848")))

# Reorder Stim concentration factor levels
samples_data_plot <- samples_data_plot %>%
  mutate(Stim_Conc = factor(Stim_Conc, levels = c("0", "1", "500")))

# Define a colour palette for cytokines
cytokine_palette <- c("Untreated" = "#666666", 
                      "IFNa" = "#d95f02", 
                      "IFNb" = "#66a61e", 
                      "IFNg" = "#e7298a", 
                      "TNF" = "#7570b3", 
                      "IL-1b" = "#e6ab02", 
                      "IL-6" = "#a6761d", 
                      "IL-23" = "#1b9e77")

# Define an alpha palette for stim concentrations
alpha_palette <- c("0" = 0.5, "1" = 1, "500" = 1)

# Setting stim conc as categorical variable for plotting purposes
samples_data_plot <- samples_data_plot %>%
  mutate(Stim_Conc = factor(Stim_Conc, levels = c("0", "1", "500")))
```

### Plotting an overview

The objective is to compare the methods (Hamilton, Hamilton 1 and Manual), and see how the sample concentrations differ between them, and how they correlate with each other. We can make a scatter plot with the concentrations from one method on the x axis, and the concentrations from another method on the y axis, and color the points by treatment or stim condition. We can also add a line of identity (y = x) to see how well the methods agree with each other.

```{r overview plot}
# Reshape the data frame to have one column for each donor's concentrations
samples_data_overview <- samples_data_plot %>%
  select(Well, Sample, Cytokine, PreTreatment, Stim, Stim_Conc, 
         Donor_ID, Timepoint, Final_Concentration_pg_mL) %>%
  summarise(
    Final_Concentration_pg_mL = mean(Final_Concentration_pg_mL, na.rm = TRUE), 
    .by = c(Sample, Cytokine, PreTreatment, Stim, Stim_Conc, Donor_ID, 
            Timepoint)
    )

samples_data_overview <- samples_data_overview %>%
  select(-c(Sample)) %>%
  pivot_wider(names_from = Donor_ID, values_from = Final_Concentration_pg_mL)

# Scatter plot comparing Hamilton and Manual methods
overview_plot <- ggplot(samples_data_overview, 
                        aes(x = `Donor_1`, 
                            y = `Donor_2`, 
                            colour = PreTreatment, 
                            shape = Stim, 
                            alpha = Stim_Conc)) +
  geom_point(size = 3) +
  labs(title = "Comparison of Donor plasma concentrations",
       x = "Donor 1 Concentration (pg/mL)",
       y = "Donor 2 Concentration (pg/mL)") +
  theme_minimal() +
  scale_color_manual(values = cytokine_palette) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  facet_wrap(~ Cytokine + Timepoint, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(overview_plot)

ggsave(overview_plot, 
       filename = paste0(output_folder, 
                         experiment_id, 
                         "_Overview_Donors_AllCytokines.pdf"
                         ), 
       width = 8, 
       height = 6
       )
```

### Boxplots comparing pretreatments for each stimulation

```{r palettes}
# Reshape the data frame to have one column for each method's concentrations
samples_data_boxplot <- samples_data_plot

# Colours for within range, above ULOQ, below LLOD
range_palette <- c("Below LLOD" = "yellow3", 
                   "Between LLOD and LLOQ" = "grey80", 
                   "Within Range" = "black", 
                   "Above ULOQ" = "red")

# Shape used to differentiate donors
shape_palette <- c("Donor_1" = 16, "Donor_2" = 17)

# Alpha palette for timepoints
timepoint_palette <- c("1h" = 0.5, "20h" = 1)
```

Boxplots for comparing timepoints, by also having position dodge for timepoints, and faceting by stim and cytokine.

```{r boxplots comparing methods}
# Position dodge boxplots, facet by stim and cytokine
boxplot_timepoint <- ggplot(samples_data_boxplot, 
                            aes(x = PreTreatment, 
                                y = Final_Concentration_pg_mL, 
                                fill = PreTreatment)) +
  geom_boxplot(outlier.shape = NA, 
               position = position_dodge(width = 0.8), 
               aes(alpha = Timepoint)) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, 
                                              dodge.width = 0.8), 
              size = 1.5, 
              aes(color = Within_Range, shape = Donor_ID, alpha = Timepoint)
              ) +
  labs(title = "Effect of pre-treatments on plasma cytokine concentrations",
       subtitle = "1h pre-treatment, comparing 1h and 20h stimulations",
       x = "Pre-treatment",
       y = "Plasma concentration (pg/mL)") +
  theme_minimal() +
  scale_fill_manual(values = cytokine_palette) +
  scale_color_manual(values = range_palette) +
  scale_alpha_manual(values = timepoint_palette) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(Cytokine ~ Stim, ncol = 3, axes = "all_y", scales = "free_y")

# save the plot
ggsave(boxplot_timepoint, 
       filename = paste0(output_folder, 
                         experiment_id, 
                         "_Boxplot_PreTreatmentsvsStims_Donors_IFNavsTNF.pdf"),
       width = 12, 
       height = 8
)
print(boxplot_timepoint)
```
